---
title: "Apply lab calibration — GRIP snowmelt head"
subtitle: "Per-node/per-NTC Steinhart–Hart (S4) vs. the standard Beta formula"
author: "Thom Laepple"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 9
    fig-height: 6
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
---

## Idea

Each NTC has an independent lab calibration (counts → temperature) measured
against an SPRT reference: a **4-parameter Steinhart–Hart** fit per sensor. Here
we apply those per-sensor coefficients to the GRIP satellite data and compare to
the **standard fixed-Beta formula** used by the field decoder so far.

Calibration chain (same `counts → R` stage as the decoder, new `R → T` stage):

$$
H = \frac{-\text{counts}\cdot 10^6}{\text{counts}-33554432}, \quad
R = \frac{H\cdot 499000}{499000-H}, \quad
\frac{1}{T} = a + b\ln R + c\ln^2 R + d\ln^3 R
$$

with per-sensor $(a,b,c,d)$ from `meta/GRIP_calibration_assignment.csv` (built by
`build_calibration_assignment.R`). **Mapping:** satellite node index `i` (1..24,
bottom→top, transmission order) = `position i` in the assignment.

```{r setup}
library(dplyr); library(tidyr); library(ggplot2); library(patchwork)

IMEI    <- "300434065508020"
fields  <- c("ntc1", "ntc2", "testSB")   # = NTC1, NTC2, NTC3
n_nodes <- 24L
spacing <- 4L
fmap    <- c(ntc1 = 1L, ntc2 = 2L, testSB = 3L)   # field -> NTC index
```

## Functions

```{r fns}
bin_field_bytes <- function(f) {
  if (f %in% c("ntc1", "ntc2", "testSB", "pressure")) return(3L)
  if (f == "gnd") return(2L); stop("Unknown field: ", f)
}
read_snowchain_bin <- function(path, fields, n_nodes) {
  raw <- readBin(path, what = "raw", n = file.info(path)$size)
  n <- length(raw); ints <- as.integer(raw); has_sep <- ints[5] == 0x7CL
  be_uint <- function(b) sum(as.integer(b) * 256^rev(seq_along(b) - 1L))
  be_sint <- function(b) { v <- be_uint(b); bits <- 8L * length(b)
    if (v >= 2^(bits - 1L)) v - 2^bits else v }
  nf <- length(fields)
  widths <- c(3L, rep(sapply(fields, bin_field_bytes), n_nodes), 2L)
  signed <- c(TRUE, rep(fields != "pressure", n_nodes), FALSE); n_val <- length(widths)
  decode_one <- function(b, w, sgn) {
    if (length(b) != w) return(NA_real_); if (sgn) be_sint(b) else be_uint(b) }
  if (has_sep) {
    sp <- which(ints == 0x7CL); sp <- sp[sp >= 5L]
    fstart <- sp + 1L; fend <- c(sp[-1] - 1L, n)
    vals <- lapply(seq_len(min(length(sp), n_val)), function(i)
      decode_one(raw[fstart[i]:fend[i]], widths[i], signed[i]))
  } else {
    pos <- 5L; vals <- vector("list", n_val)
    for (i in seq_len(n_val)) { w <- widths[i]
      if (w == 3L && pos + 3L <= n && all(ints[pos:(pos + 3L)] == c(1L, 0L, 0L, 0L))) {
        vals[[i]] <- NA_real_; pos <- pos + 4L
      } else if (i == n_val && pos + w - 1L > n && pos <= n) {
        vals[[i]] <- be_uint(raw[pos:n]) * 256L^(w - (n - pos + 1L)); pos <- n + 1L
      } else { vals[[i]] <- decode_one(raw[pos:(pos + w - 1L)], w, signed[i]); pos <- pos + w }
    }
  }
  nv <- vals[2:(1L + n_nodes * nf)]
  list(time_unix = be_uint(raw[1:4]),
       nodes = lapply(seq_len(n_nodes), function(i)
         setNames(nv[((i - 1L) * nf + 1L):(i * nf)], fields)))
}

# counts -> resistance (decoder stage), and the two R -> T conversions.
NTCcounts2R    <- function(counts) {
  H <- (-counts * 1e6) / (counts - 33554432); (H * 499000) / (499000 - H) }
NTCcounts2temp <- function(counts) {           # standard fixed-Beta
  R <- NTCcounts2R(counts); 1 / ((1 / 3380) * log(R / 10000) + 1 / 298.15) - 273.15 }
S4_predict_T_C <- function(R, co) {            # per-sensor Steinhart-Hart
  lR <- log(R); 1 / (co[1] + co[2] * lR + co[3] * lR^2 + co[4] * lR^3) - 273.15 }
```

## Apply

```{r apply}
asg <- read.csv("../meta/GRIP_calibration_assignment.csv")

files <- Sys.glob(sprintf("../testdata/%s-*.bin", IMEI))
files <- files[!grepl(sprintf("%s-2\\.bin$", IMEI), files)]   # head not yet erected
all   <- lapply(files, read_snowchain_bin, fields = fields, n_nodes = n_nodes)
ord   <- order(sapply(all, `[[`, "time_unix")); all <- all[ord]
times <- as.POSIXct(sapply(all, `[[`, "time_unix"), origin = "1970-01-01", tz = "UTC")
n_msg <- length(all)

# Lookup of coefficients keyed by (position == node, ntc).
co_of <- function(node, k) {
  r <- asg[asg$position == node & asg$ntc == k, c("a", "b", "c", "d")]
  if (nrow(r) == 1) as.numeric(r) else rep(NA_real_, 4)
}

rows <- list()
for (m in seq_len(n_msg)) for (i in seq_len(n_nodes)) for (f in names(fmap)) {
  k   <- fmap[[f]]; cnt <- all[[m]]$nodes[[i]][[f]]
  R   <- if (is.na(cnt)) NA_real_ else NTCcounts2R(cnt)
  Ts  <- if (is.na(cnt)) NA_real_ else NTCcounts2temp(cnt)
  Tc  <- if (is.na(cnt)) NA_real_ else S4_predict_T_C(R, co_of(i, k))
  rows[[length(rows) + 1L]] <- data.frame(
    msg = m, time = times[m], node = i, ntc = k,
    counts = cnt, T_std = Ts, T_cal = Tc, diff = Tc - Ts)
}
cal <- bind_rows(rows) |>
  mutate(sensor   = factor(ntc, 1:3, c("NTC1", "NTC2", "NTC3")),
         hoehe_cm = (node - 1L) * 3L * spacing + ntc * spacing)
```

## Difference: lab vs. standard

```{r diff-table}
cal |> group_by(sensor) |>
  summarise(`mean Δ (°C)` = mean(diff, na.rm = TRUE),
            `sd Δ`        = sd(diff,   na.rm = TRUE),
            `min Δ`       = min(diff,  na.rm = TRUE),
            `max Δ`       = max(diff,  na.rm = TRUE),
            n = sum(!is.na(diff)), .groups = "drop") |>
  knitr::kable(digits = 3, caption = "T_cal − T_std (lab minus standard) over all messages")
```

The lab calibration is a remarkably **uniform cold offset** relative to the fixed
Beta formula (note the tiny SD). The high-`a` coefficients flagged earlier are
**not** outliers in temperature — `b,c,d` compensate:

```{r diff-vs-a, fig.height=4, fig.width=9}
cal_a <- cal |> filter(msg == n_msg) |>
  left_join(asg |> transmute(node = position, ntc, a, a_suspect),
            by = c("node", "ntc"))
ggplot(cal_a, aes(x = a, y = diff, colour = a_suspect)) +
  geom_point(size = 2) +
  scale_colour_manual(values = c(`FALSE` = "#34495e", `TRUE` = "#e74c3c"),
                      labels = c("normal", "a_suspect (z>4)")) +
  labs(title = "Calibration shift vs. the 'a' coefficient (newest message)",
       subtitle = "flagged high-a sensors give normal temperature shifts → fits are fine",
       x = "a coefficient", y = "T_cal − T_std (°C)", colour = NULL) +
  theme_bw(base_size = 13) + theme(legend.position = "bottom")
```

```{r diff-profile, fig.height=5, fig.width=8}
cal |> filter(msg == n_msg) |>
  ggplot(aes(x = diff, y = hoehe_cm, colour = sensor)) +
  geom_path(aes(group = sensor), alpha = 0.5) + geom_point(size = 2) +
  scale_colour_manual(values = c(NTC1 = "#2980b9", NTC2 = "#e74c3c", NTC3 = "#27ae60")) +
  labs(title = "Per-node calibration shift (newest message)",
       subtitle = sprintf("%s UTC", format(times[n_msg], "%Y-%m-%d %H:%M")),
       x = "T_cal − T_std (°C)", y = "Höhe (cm)", colour = NULL) +
  theme_bw(base_size = 13) + theme(legend.position = "bottom")
```

## Calibrated field result

Newest profile — standard (grey) vs. lab-calibrated (coloured):

```{r profile, fig.height=7, fig.width=7}
prof <- cal |> filter(msg == n_msg) |>
  select(hoehe_cm, sensor, T_std, T_cal) |>
  pivot_longer(c(T_std, T_cal), names_to = "kind", values_to = "temp_C") |>
  filter(is.finite(temp_C)) |>
  mutate(kind = recode(kind, T_std = "Standard (Beta)", T_cal = "Lab-kalibriert (S4)"))
ggplot(prof, aes(x = temp_C, y = hoehe_cm, colour = sensor, linetype = kind, shape = kind)) +
  geom_path(aes(group = interaction(sensor, kind)), alpha = 0.6) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = c(NTC1 = "#2980b9", NTC2 = "#e74c3c", NTC3 = "#27ae60")) +
  scale_linetype_manual(values = c("Standard (Beta)" = "dotted", "Lab-kalibriert (S4)" = "solid")) +
  scale_shape_manual(values = c("Standard (Beta)" = 1, "Lab-kalibriert (S4)" = 16)) +
  labs(title = "Profil — Standard vs. Lab-Kalibration",
       subtitle = sprintf("%s UTC | IMEI %s", format(times[n_msg], "%Y-%m-%d %H:%M"), IMEI),
       x = "Temperatur (°C)", y = "Höhe (cm)", colour = NULL, linetype = NULL, shape = NULL) +
  theme_bw(base_size = 13) + theme(legend.position = "bottom")
```

Calibrated time series (heatmap + lines), `eval` only with >1 message:

```{r cal-series, eval=(n_msg > 1), fig.height=11, fig.width=8}
ser <- cal |> filter(is.finite(T_cal))
tlevels <- format(sort(unique(ser$time)), "%m-%d %H:%M")
ser <- ser |> mutate(tlab = factor(format(time, "%m-%d %H:%M"), levels = tlevels))
max_h <- n_nodes * 3L * spacing

p_heat <- ggplot(ser, aes(x = tlab, y = hoehe_cm, fill = T_cal)) +
  geom_tile() +
  scale_fill_distiller(palette = "RdYlBu", direction = -1, name = "T (°C)") +
  scale_y_continuous("Höhe (cm)", limits = c(0, max_h),
                     breaks = seq(0, max_h, by = 8L), expand = c(0, 0)) +
  labs(title = "Zeitlicher Verlauf — lab-kalibriert", x = NULL,
       subtitle = sprintf("%d Messungen | IMEI %s", n_msg, IMEI)) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())

p_line <- ggplot(ser, aes(x = time, y = T_cal,
                          group = interaction(node, sensor), colour = hoehe_cm)) +
  geom_line(alpha = 0.5, linewidth = 0.4) + geom_point(size = 1) +
  scale_colour_viridis_c(name = "Höhe (cm)") +
  labs(x = "Zeit (UTC)", y = "Temperatur (°C)") +
  theme_bw(base_size = 13)

p_heat / p_line + plot_layout(heights = c(3, 2))
```

## Anomaly (profile mean removed): uncalibrated vs. calibrated

At each timepoint the mean over all heights is subtracted, leaving the vertical
structure. Shown for the uncalibrated (standard Beta) and lab-calibrated (S4)
temperatures.

```{r anom-build, eval=(n_msg > 1)}
anom <- cal |>
  select(time, node, ntc, sensor, hoehe_cm, T_std, T_cal) |>
  pivot_longer(c(T_std, T_cal), names_to = "kind", values_to = "temp_C") |>
  filter(is.finite(temp_C)) |>
  mutate(kind = recode(kind, T_std = "uncalibrated (Beta)", T_cal = "lab-calibrated (S4)"),
         kind = factor(kind, levels = c("uncalibrated (Beta)", "lab-calibrated (S4)"))) |>
  group_by(time, kind) |>
  mutate(anom = temp_C - mean(temp_C, na.rm = TRUE)) |>
  ungroup()
tlevels <- format(sort(unique(anom$time)), "%m-%d %H:%M")
anom  <- anom |> mutate(tlab = factor(format(time, "%m-%d %H:%M"), levels = tlevels))
alim  <- max(abs(anom$anom), na.rm = TRUE)
max_h <- n_nodes * 3L * spacing
```

```{r anom-heat, eval=(n_msg > 1), fig.height=6, fig.width=10}
ggplot(anom, aes(x = tlab, y = hoehe_cm, fill = anom)) +
  geom_tile() +
  facet_wrap(~ kind) +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, limits = c(-alim, alim), name = "ΔT (°C)") +
  scale_y_continuous("Höhe (cm)", limits = c(0, max_h),
                     breaks = seq(0, max_h, by = 8L), expand = c(0, 0)) +
  labs(title = "Anomaly — profile mean removed per timepoint", x = NULL) +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
```

```{r anom-line, eval=(n_msg > 1), fig.height=5, fig.width=10}
ggplot(anom, aes(x = time, y = anom, group = interaction(node, sensor), colour = hoehe_cm)) +
  geom_hline(yintercept = 0, linewidth = 0.4, colour = "grey50") +
  geom_line(alpha = 0.5, linewidth = 0.4) + geom_point(size = 0.8) +
  facet_wrap(~ kind) +
  scale_colour_viridis_c(name = "Höhe (cm)") +
  labs(x = "Zeit (UTC)", y = "T − profile mean (°C)") +
  theme_bw(base_size = 13)
```

## Save

```{r save}
out <- cal |>
  select(msg, time, node, ntc, sensor, hoehe_cm, counts, T_std, T_cal, diff) |>
  arrange(time, node, ntc)
outfile <- sprintf("data/labcalibrated_imei%s_%s.csv", IMEI,
                   format(times[n_msg], "%Y%m%d_%H%M%S"))
write.csv(out, outfile, row.names = FALSE)
cat("Saved:", outfile, "\n")
```
