---
title: "NTC2 sensitivity: calibration slope vs. raw counts"
author: "Thom Laepple"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    fig-width: 9
    fig-height: 5
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
---

In the field data the NTC2 channel shows a ~2.4 % larger diurnal temperature
amplitude than NTC1/NTC3 (per node, paired). The temperature amplitude is the
product of two factors:

$$\Delta T = \frac{dT}{d\,\text{counts}} \times \Delta\text{counts}.$$

This note measures both factors separately: the calibration slope
$dT/d\text{counts}$ from the lab coefficients, and the raw counts amplitude from
the satellite data.

```{r setup}
suppressMessages({library(dplyr); library(tidyr); library(ggplot2)})

# counts -> resistance (decoder stage) and the 4-parameter Steinhart-Hart R -> T.
NTCcounts2R <- function(c) { H <- (-c * 1e6) / (c - 33554432); (H * 499000) / (499000 - H) }
S4_T_C      <- function(R, co) { lR <- log(R); 1 / (co[1] + co[2]*lR + co[3]*lR^2 + co[4]*lR^3) - 273.15 }
```

## 1. Calibration slope $dT/d\text{counts}$ at −5 °C (all SM1 + SM2 sensors)

```{r calib-slope}
read_sm <- function(f, src) {
  m <- read.csv(f, check.names = FALSE, row.names = 1)       # rows a,b,c,d
  do.call(rbind, lapply(colnames(m), function(cn) {
    p <- strsplit(cn, "_")[[1]]
    data.frame(source = src, node = as.integer(sub("N", "", p[1])),
               ntc = as.integer(sub("NTC", "", p[2])),
               a = m["a", cn], b = m["b", cn], c = m["c", cn], d = m["d", cn])
  }))
}
co <- rbind(read_sm("../meta/SM1_CalibrationCoefficients.csv", "SM1"),
            read_sm("../meta/SM2_CalibrationCoefficients.csv", "SM2"))
co <- co[complete.cases(co[, c("a", "b", "c", "d")]), ]

T0 <- -5
slope_at <- function(a, b, c, d) {
  cf <- c(a, b, c, d)
  Tc <- function(cc) S4_T_C(NTCcounts2R(cc), cf)
  f  <- function(cc) Tc(cc) - T0
  if (sign(f(7e5)) == sign(f(1.5e6))) return(c(NA, NA))
  c0 <- uniroot(f, c(7e5, 1.5e6), tol = 1)$root
  c(c0, (Tc(c0 + 200) - Tc(c0 - 200)) / 400)            # counts@-5C , dT/dcounts
}
r <- t(mapply(slope_at, co$a, co$b, co$c, co$d))
co$counts_at_m5 <- r[, 1]
co$mK_per_kcount <- r[, 2] * 1e6                        # mK per 1000 counts
ok <- co |> filter(is.finite(mK_per_kcount), counts_at_m5 > 8e5, counts_at_m5 < 1.3e6)

ok |> group_by(NTC = ntc) |>
  summarise(n = n(),
            `slope mean (mK/kcount)`   = mean(mK_per_kcount),
            `slope sd`                 = sd(mK_per_kcount),
            `counts@-5C mean`          = mean(counts_at_m5), .groups = "drop") |>
  knitr::kable(digits = 3, caption = sprintf("Local counts->T slope at -5 C (%d of %d sensors)", nrow(ok), nrow(co)))
```

```{r calib-test}
s <- split(ok$mK_per_kcount, ok$ntc)
data.frame(
  comparison = c("NTC2 vs NTC1", "NTC2 vs NTC3"),
  `diff (mK/kcount)` = c(mean(s[["2"]]) - mean(s[["1"]]), mean(s[["2"]]) - mean(s[["3"]])),
  `p (Welch)` = c(t.test(s[["2"]], s[["1"]])$p.value, t.test(s[["2"]], s[["3"]])$p.value),
  `ratio NTC2 / mean(NTC1,NTC3)` = mean(s[["2"]]) / mean(c(s[["1"]], s[["3"]])),
  check.names = FALSE
) |> knitr::kable(digits = 4)
```

```{r calib-plot, fig.height=4}
ggplot(ok, aes(factor(ntc), mK_per_kcount)) +
  geom_boxplot(outlier.size = 0.6, fill = "#d6eaf8") +
  labs(x = "NTC", y = "dT/dcounts at -5 C (mK / 1000 counts)") +
  theme_bw(base_size = 13)
```

## 2. Raw counts amplitude per NTC (field data, no calibration)

```{r counts-amp}
bin_field_bytes <- function(f) { if (f %in% c("ntc1","ntc2","testSB","pressure")) 3L else if (f=="gnd") 2L else stop() }
read_bin <- function(path, fields, n_nodes) {
  raw <- readBin(path, "raw", file.info(path)$size); n <- length(raw); ints <- as.integer(raw)
  has_sep <- ints[5] == 0x7CL
  be <- function(b) sum(as.integer(b) * 256^rev(seq_along(b) - 1L))
  bs <- function(b) { v <- be(b); bt <- 8L*length(b); if (v >= 2^(bt-1L)) v - 2^bt else v }
  nf <- length(fields); w <- c(3L, rep(sapply(fields, bin_field_bytes), n_nodes), 2L)
  sg <- c(TRUE, rep(fields != "pressure", n_nodes), FALSE); nv <- length(w)
  d1 <- function(b, ww, s) { if (length(b) != ww) NA_real_ else if (s) bs(b) else be(b) }
  if (has_sep) {
    sp <- which(ints == 0x7CL); sp <- sp[sp >= 5L]; fs <- sp + 1L; fe <- c(sp[-1] - 1L, n)
    vl <- lapply(seq_len(min(length(sp), nv)), function(i) d1(raw[fs[i]:fe[i]], w[i], sg[i]))
  } else {
    p <- 5L; vl <- vector("list", nv)
    for (i in seq_len(nv)) { ww <- w[i]
      if (ww == 3L && p + 3L <= n && all(ints[p:(p+3L)] == c(1,0,0,0))) { vl[[i]] <- NA_real_; p <- p + 4L }
      else if (i == nv && p + ww - 1L > n && p <= n) { vl[[i]] <- be(raw[p:n]) * 256L^(ww - (n-p+1L)); p <- n + 1L }
      else { vl[[i]] <- d1(raw[p:(p+ww-1L)], ww, sg[i]); p <- p + ww } }
  }
  ndv <- vl[2:(1L + n_nodes*nf)]
  list(nodes = lapply(seq_len(n_nodes), function(i) setNames(ndv[((i-1L)*nf+1L):(i*nf)], fields)),
       time = be(raw[1:4]))
}
fields <- c("ntc1", "ntc2", "testSB"); n_nodes <- 24L
files <- sort(Sys.glob("../testdata/300434065508020-*.bin"))
files <- files[!grepl("-2\\.bin$", files)]                # head not yet erected
all <- lapply(files, read_bin, fields = fields, n_nodes = n_nodes)
all <- all[order(sapply(all, `[[`, "time"))]
cmat <- function(f) sapply(all, function(b) sapply(b$nodes, function(nd) { v <- nd[[f]]; if (is.null(v)) NA_real_ else as.numeric(v) }))
amp <- function(M) apply(M, 1, function(x) { x <- x[is.finite(x)]; if (length(x) < 2) NA else diff(range(x)) })
a1 <- amp(cmat("ntc1")); a2 <- amp(cmat("ntc2")); a3 <- amp(cmat("testSB")); ref <- (a1 + a3) / 2

data.frame(
  NTC = c("NTC1", "NTC2", "NTC3"),
  `counts amplitude (mean over nodes)` = c(mean(a1, na.rm = TRUE), mean(a2, na.rm = TRUE), mean(a3, na.rm = TRUE)),
  check.names = FALSE
) |> knitr::kable(digits = 0, caption = sprintf("Raw counts amplitude (max-min over %d messages)", length(all)))

data.frame(
  metric = c("NTC2 / mean(NTC1,NTC3)", "NTC2 strongest in N nodes", "paired t-test p"),
  value  = c(sprintf("%.4f", mean(a2, na.rm = TRUE) / mean(c(a1, a3), na.rm = TRUE)),
             sprintf("%d / %d", sum(a2 > a1 & a2 > a3, na.rm = TRUE), sum(is.finite(a2 - ref))),
             sprintf("%.2g", t.test(a2, ref, paired = TRUE)$p.value))
) |> knitr::kable()
```

## 3. Node-assignment validation

```{r val-setup}
NTCstd <- function(c) { R <- NTCcounts2R(c); 1 / ((1/3380)*log(R/10000) + 1/298.15) - 273.15 }
asg <- read.csv("../meta/GRIP_calibration_assignment.csv")
pos <- asg |> distinct(position, node) |> arrange(position)
co_of <- function(node, k) {
  r <- asg[asg$node == node & asg$ntc == k, c("a","b","c","d")]
  if (nrow(r) == 1) as.numeric(r) else rep(NA_real_, 4) }
fmap <- c(ntc1 = 1L, ntc2 = 2L, testSB = 3L); nM <- length(all)
times <- as.POSIXct(sapply(all, `[[`, "time"), origin = "1970-01-01", tz = "UTC")
hgt   <- outer(seq_len(n_nodes), 1:3, function(nd, k) (nd - 1) * 12 + k * 4)
ntc_f <- function(x) factor(x, 1:3, c("NTC1", "NTC2", "NTC3"))
ntc_pal  <- c(NTC1 = "#2980b9", NTC2 = "#e74c3c", NTC3 = "#27ae60")
kind_shp <- c(`uncalibrated (Beta)` = 1,        `lab-calibrated (S4)` = 16)
kind_lty <- c(`uncalibrated (Beta)` = "dashed", `lab-calibrated (S4)` = "solid")

Cnt <- array(NA_real_, c(n_nodes, 3, nM))
for (m in seq_len(nM)) for (i in seq_len(n_nodes)) for (f in names(fmap)) {
  v <- all[[m]]$nodes[[i]][[f]]; Cnt[i, fmap[[f]], m] <- if (is.null(v)) NA else as.numeric(v)
}
Tfield <- function(mode, nfp = NULL) {
  T <- array(NA_real_, c(n_nodes, 3, nM))
  for (i in seq_len(n_nodes)) { nd <- if (is.null(nfp)) pos$node[i] else nfp[i]
    for (k in 1:3) { co <- co_of(nd, k)
      for (m in seq_len(nM)) { cc <- Cnt[i, k, m]; if (is.na(cc)) next
        T[i, k, m] <- if (mode == "std") NTCstd(cc) else S4_T_C(NTCcounts2R(cc), co) } } }
  T
}
Ts <- Tfield("std"); Tc <- Tfield("cal")
# Use the quietest (smoothest) profile: the message with the lowest per-NTC
# neighbour-difference variance.
msg_rough <- sapply(seq_len(nM), function(m)
  mean(sapply(1:3, function(k) var(diff(Ts[, k, m]), na.rm = TRUE)), na.rm = TRUE))
m_ex <- which.min(msg_rough)
ttl  <- format(times[m_ex], "%Y-%m-%d %H:%M")
```

### Example profiles (quietest profile)

The timepoint with the smoothest profile (lowest per-NTC neighbour-difference
variance) is used for the example plots.

Step 1 — absolute temperatures, all NTCs. The within-node NTC pattern (sawtooth)
and the calibration shift are visible:

```{r ex-absolute, fig.height=4, fig.width=9}
df_abs <- rbind(
  data.frame(node = rep(seq_len(n_nodes), 3), ntc = rep(1:3, each = n_nodes),
             hoehe = as.vector(hgt), temp = as.vector(Ts[, , m_ex]), kind = "uncalibrated (Beta)"),
  data.frame(node = rep(seq_len(n_nodes), 3), ntc = rep(1:3, each = n_nodes),
             hoehe = as.vector(hgt), temp = as.vector(Tc[, , m_ex]), kind = "lab-calibrated (S4)"))
ggplot(df_abs, aes(hoehe, temp)) +
  geom_line(aes(group = kind, linetype = kind), colour = "grey65", linewidth = 0.4) +
  geom_point(aes(colour = ntc_f(ntc), shape = kind), size = 1.9) +
  scale_colour_manual(values = ntc_pal) +
  scale_shape_manual(values = kind_shp) + scale_linetype_manual(values = kind_lty) +
  labs(title = sprintf("Absolute temperature, all NTCs — %s UTC", ttl),
       x = "Höhe (cm)", y = "Temperature (°C)", colour = "NTC", shape = NULL, linetype = NULL) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

Step 2 — anomaly (profile mean removed per kind), all NTCs. The sawtooth remains;
uncalibrated and calibrated overlap:

```{r ex-anom-all, fig.height=4, fig.width=9}
df_an <- df_abs |> group_by(kind) |> mutate(anom = temp - mean(temp, na.rm = TRUE)) |> ungroup()
ggplot(df_an, aes(hoehe, anom)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_line(aes(group = kind, linetype = kind), colour = "grey65", linewidth = 0.4) +
  geom_point(aes(colour = ntc_f(ntc), shape = kind), size = 1.9) +
  scale_colour_manual(values = ntc_pal) +
  scale_shape_manual(values = kind_shp) + scale_linetype_manual(values = kind_lty) +
  labs(title = "Anomaly, all NTCs", x = "Höhe (cm)", y = "T − profile mean (°C)",
       colour = "NTC", shape = NULL, linetype = NULL) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

Sawtooth count — within each node, position of NTC2 in its NTC1/NTC2/NTC3 packet
(`d2 = NTC2 − mean(NTC1,NTC3)`). Chance for coldest/warmest ≈ 1/3:

```{r ex-sawtooth}
saw <- do.call(rbind, lapply(seq_len(nM), function(m) {
  d2 <- Ts[, 2, m] - (Ts[, 1, m] + Ts[, 3, m]) / 2
  is2 <- function(fn) sum(sapply(seq_len(n_nodes), function(i) {
    x <- Ts[i, , m]; if (any(is.na(x))) NA else fn(x) == 2 }), na.rm = TRUE)
  n <- sum(is.finite(d2))
  data.frame(message = format(times[m], "%m-%d %H:%M"),
             `mean T (°C)`        = round(mean(Ts[, , m], na.rm = TRUE), 1),
             `mean d2 (°C)`       = round(mean(d2, na.rm = TRUE), 3),
             `NTC2 below mid`     = sprintf("%d/%d", sum(d2 < 0, na.rm = TRUE), n),
             `NTC2 coldest`       = sprintf("%d/%d", is2(which.min), n),
             `NTC2 warmest`       = sprintf("%d/%d", is2(which.max), n),
             check.names = FALSE)
}))
knitr::kable(saw, caption = "NTC2 within its node packet. NTC2 is systematically coldest when cold and warmest when warm; the sawtooth flips sign at the crossover (~-12 °C), where the quietest profile sits.")
```

Same anomaly plot for the message with the **clearest sawtooth** (largest mean
|d2|), with and without calibration:

```{r ex-anom-saw, fig.height=4, fig.width=9}
md2   <- sapply(seq_len(nM), function(m) mean(Ts[, 2, m] - (Ts[, 1, m] + Ts[, 3, m]) / 2, na.rm = TRUE))
m_saw <- which.max(abs(md2))
df_abs_s <- rbind(
  data.frame(node = rep(seq_len(n_nodes), 3), ntc = rep(1:3, each = n_nodes),
             hoehe = as.vector(hgt), temp = as.vector(Ts[, , m_saw]), kind = "uncalibrated (Beta)"),
  data.frame(node = rep(seq_len(n_nodes), 3), ntc = rep(1:3, each = n_nodes),
             hoehe = as.vector(hgt), temp = as.vector(Tc[, , m_saw]), kind = "lab-calibrated (S4)"))
df_an_s <- df_abs_s |> group_by(kind) |> mutate(anom = temp - mean(temp, na.rm = TRUE)) |> ungroup()
ggplot(df_an_s, aes(hoehe, anom)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_line(aes(group = kind, linetype = kind), colour = "grey65", linewidth = 0.4) +
  geom_point(aes(colour = ntc_f(ntc), shape = kind), size = 1.9) +
  scale_colour_manual(values = ntc_pal) +
  scale_shape_manual(values = kind_shp) + scale_linetype_manual(values = kind_lty) +
  labs(title = sprintf("Anomaly, all NTCs — clearest sawtooth (%s UTC)",
                       format(times[m_saw], "%Y-%m-%d %H:%M")),
       x = "Höhe (cm)", y = "T − profile mean (°C)", colour = "NTC", shape = NULL, linetype = NULL) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

Step 3 — same anomaly split by NTC:

```{r ex-anom-ntc, fig.height=4, fig.width=10}
ggplot(df_an, aes(node, anom, colour = ntc_f(ntc))) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_line(aes(linetype = kind)) + geom_point(aes(shape = kind), size = 1.6) +
  facet_wrap(~ ntc_f(ntc)) +
  scale_colour_manual(values = ntc_pal, guide = "none") +
  scale_shape_manual(values = kind_shp) + scale_linetype_manual(values = kind_lty) +
  labs(title = "Anomaly by NTC", x = "Node (bottom → top)", y = "T − profile mean (°C)",
       shape = NULL, linetype = NULL) +
  theme_bw(base_size = 12) + theme(legend.position = "bottom")
```

### Roughness metric: variance of neighbour differences

Roughness = variance of node-to-node differences within each NTC, averaged over
the 3 NTCs and the messages. Compared for uncalibrated, correct, and random
node→coefficient assignments:

```{r metric, fig.height=4}
rough <- function(T) mean(sapply(seq_len(nM), function(m)
  mean(sapply(1:3, function(k) var(diff(T[, k, m]), na.rm = TRUE)), na.rm = TRUE)))
r_uncal <- rough(Ts); r_corr <- rough(Tc)
set.seed(1)
r_rand <- replicate(200, rough(Tfield("cal", pos$node[sample(n_nodes)])))

knitr::kable(data.frame(
  assignment = c("uncalibrated (Beta)", "correct", "random (200x) mean"),
  `neighbour-diff variance (C^2)` = c(r_uncal, r_corr, mean(r_rand)),
  check.names = FALSE), digits = 4,
  caption = sprintf("Correct assignment is smoother than %d%% of random draws.",
                    round(100 * mean(r_rand > r_corr))))

hist(r_rand, breaks = 20, col = "grey85", border = "white",
     main = "Random node assignments (200x)", xlab = "neighbour-diff variance (per NTC)",
     xlim = range(c(r_rand, r_corr, r_uncal)))
abline(v = r_corr,  col = "#c0392b", lwd = 2)
abline(v = r_uncal, col = "#2980b9", lwd = 2, lty = 2)
legend("topright", c("correct", "uncalibrated"),
       col = c("#c0392b", "#2980b9"), lwd = 2, lty = c(1, 2), bty = "n")
```

Correct vs. 20 random assignments, per-NTC profiles (quietest profile):

```{r ex-random, fig.height=4, fig.width=10}
per_ntc <- function(T) do.call(rbind, lapply(1:3, function(k) {
  x <- T[, k, m_ex]; data.frame(node = seq_len(n_nodes), ntc = k, anom = x - mean(x, na.rm = TRUE)) }))
set.seed(5)
df_rnd  <- do.call(rbind, lapply(1:20, function(j)
  transform(per_ntc(Tfield("cal", pos$node[sample(n_nodes)])), draw = j)))
df_corr <- per_ntc(Tc)
df_unc  <- per_ntc(Ts)
ggplot() +
  geom_hline(yintercept = 0, colour = "grey85", linewidth = 0.3) +
  geom_line(data = df_rnd, aes(node, anom, group = interaction(ntc, draw)),
            colour = "grey78", linewidth = 0.3, alpha = 0.7) +
  geom_line(data = df_unc, aes(node, anom), colour = "black", linewidth = 0.6) +
  geom_line(data = df_corr, aes(node, anom, colour = ntc_f(ntc)), linewidth = 0.8) +
  geom_point(data = df_corr, aes(node, anom, colour = ntc_f(ntc)), size = 1.2) +
  facet_wrap(~ ntc_f(ntc)) +
  scale_colour_manual(values = ntc_pal, guide = "none") +
  labs(title = sprintf("Correct (coloured) vs. 20 random assignments — %s UTC", ttl),
       subtitle = "black: uncalibrated (Beta)",
       x = "Node (bottom → top)", y = "per-NTC anomaly (°C)") +
  theme_bw(base_size = 12)
```

## Summary

```{r summary}
data.frame(
  factor = c("dT/dcounts at -5 C (calibration)", "counts amplitude (field measurement)", "= temperature amplitude (field)"),
  `NTC2 vs NTC1/NTC3` = c(sprintf("x %.4f", mean(s[["2"]]) / mean(c(s[["1"]], s[["3"]]))),
                          sprintf("x %.4f", mean(a2, na.rm = TRUE) / mean(c(a1, a3), na.rm = TRUE)),
                          "x 1.024"),
  source = c("SM1+SM2 coefficients", "raw satellite counts", "decoded temperatures"),
  check.names = FALSE
) |> knitr::kable(caption = "The NTC1/2/3 difference is in the counts, not in the counts->temperature calibration.")
```
