---
title: "head03 — Atmosphere & Firn Sensors (Kohnen 2026)"
subtitle: "Modem 231709 | NTC thermistors + pressure sensors | Decoded from Rockblock SBD"
author: "Thom Laepple"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 12
    fig-height: 6
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(patchwork)

d <- read_csv("./data/decoded_head03_231709_combined.csv", show_col_types = FALSE) |>
  mutate(time_utc = as_datetime(time_utc))

# Payloads now come from the Cloudloop API for the WHOLE record (see
# fetch_antarctic_cloudloop.R + decode_combined.R): every node n1–n35 + battery is
# valid throughout. The former 128-B web-export truncation (which invalidated
# n12+ after 2026-03-06) NO LONGER APPLIES. CUTOFF_FULL now just marks the source
# switch (old Rockblock export → Cloudloop API) for reference lines.
CUTOFF_FULL <- as_datetime("2026-03-06 00:00:00", tz = "UTC")

d <- d |> mutate(full_payload = TRUE)   # full payload everywhere → no masking
```

## Overview

Modem **231709** (head03) sends **two messages per transmission** (~09:03 and ~21:03 UTC):
a 340-byte primary message and a 54-byte continuation message that must be concatenated
before decoding. The chain carries NTC thermistors and pressure sensors.

**Data coverage:** full payload (all **n1–n35** + battery) across the whole record.
Earlier the Rockblock/Cloudloop **web CSV** export truncated payloads at 128 B, so
only n1–n11 survived after 6 Mar; the data now come from the **Cloudloop API**
(`fetch_antarctic_cloudloop.R` → `decode_combined.R`), which returns the full
340 + 54 byte message, so all nodes are valid end-to-end.

| Period | Date range | Nodes decoded | Source |
|--------|-----------|---------------|--------|
| Old export | 18 Jan – 5 Mar 2026 | n1–n35 + battery | Old Rockblock export (untruncated) |
| Cloudloop API | 6 Mar 2026 – present | n1–n35 + battery | Cloudloop API (full payload) |

**Total:** `r nrow(d)` transmissions,
`r format(min(d$time_utc), "%d %b %Y")` – `r format(max(d$time_utc), "%d %b %Y")`

---

## Temperature time series

```{r temp-overview}
ntc_cols <- grep("\\.ntc1_temp_C$", names(d), value = TRUE)
nodes_present <- as.integer(sub("n(\\d+)\\..*", "\\1",
                                 ntc_cols[sapply(ntc_cols, function(c) !all(is.na(d[[c]])))]))

ntc_long <- d |>
  select(time_utc, full_payload, all_of(ntc_cols)) |>
  pivot_longer(-c(time_utc, full_payload), names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor)),
         layer = case_when(node == 1  ~ "n1 (atmosphere)",
                           node <= 11 ~ "n2-11 (full period)",
                           TRUE       ~ "n12-35 (old period only)"))

ggplot(ntc_long |> filter(!is.na(temp_C), temp_C > -70),
       aes(time_utc, temp_C, colour = node, group = node)) +
  geom_line(linewidth = 0.5, alpha = 0.7) +
  scale_colour_viridis_c(option = "turbo", name = "Node") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  geom_vline(xintercept = CUTOFF_FULL, linetype = "dashed", colour = "grey50") +
  annotate("text", x = CUTOFF_FULL, y = -25, label = " Cloudloop API\n from here",
           hjust = 0, size = 3, colour = "grey40") +
  labs(title = "head03 (231709): all NTC temperatures",
       subtitle = "n1 (red) = atmosphere  |  subsurface nodes show seasonal cooling  |  dashed = source switch (web export → API)",
       x = NULL, y = "Temperature (°C)") +
  theme_minimal(base_size = 13)
```

### Atmosphere (n1) vs firn sensors

```{r temp-split}
p_atm <- ggplot(ntc_long |> filter(node == 1, !is.na(temp_C)),
                aes(time_utc, temp_C)) +
  geom_line(colour = "firebrick", linewidth = 0.9) +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(title = "n1 — atmosphere", x = NULL, y = "°C") +
  theme_minimal(base_size = 12)

p_firn <- ggplot(ntc_long |> filter(node > 1, node <= 11, !is.na(temp_C)),
                 aes(time_utc, temp_C, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.7, alpha = 0.85) +
  scale_colour_viridis_d(option = "viridis", name = "Node") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(title = "n2–n11 — near-surface firn (full period)", x = NULL, y = "°C") +
  theme_minimal(base_size = 12)

p_deep <- ggplot(ntc_long |> filter(node > 11, !is.na(temp_C), temp_C > -70),
                 aes(time_utc, temp_C, colour = node, group = node)) +
  geom_line(linewidth = 0.6, alpha = 0.8) +
  scale_colour_viridis_c(option = "plasma", name = "Node") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(title = "n12–n35 — deeper sensors (old period only, Jan–Mar)",
       x = NULL, y = "°C") +
  theme_minimal(base_size = 12)

p_atm / p_firn / p_deep
```

---

## Pressure sensors

The pressure values for all sensors should reflect the local air/firn overburden
pressure at ~660–705 hPa (Kohnen station, ~2892 m a.s.l.).

```{r pressure-setup}
p_cols <- grep("\\.pressure_hPa$", names(d), value = TRUE)

# Build long table; pressure already NA'd for artefact nodes in new period (setup chunk)
pres_long <- d |>
  select(time_utc, full_payload, all_of(p_cols)) |>
  pivot_longer(-c(time_utc, full_payload), names_to = "sensor", values_to = "hPa") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  filter(!is.na(hPa), hPa > 600, hPa < 800)  # keep only plausible range
```

### All sensors — time series

```{r pressure-timeseries, fig.height=7}
ggplot(pres_long, aes(time_utc, hPa, colour = node, group = node)) +
  geom_line(linewidth = 0.5, alpha = 0.75) +
  scale_colour_viridis_c(option = "plasma", direction = -1, name = "Node") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  geom_vline(xintercept = CUTOFF_FULL, linetype = "dashed", colour = "grey50") +
  annotate("text", x = CUTOFF_FULL, y = 800, label = " Cloudloop API\n from here",
           hjust = 0, size = 3, colour = "grey40") +
  labs(title = "head03: pressure all nodes (plausible values 600–800 hPa only)",
       subtitle = "Seasonal pressure increase Jan → Jul visible; n1 most variable; dashed = source switch (web export → API)",
       x = NULL, y = "Pressure (hPa)") +
  theme_minimal(base_size = 13)
```

### Sensor quality assessment

```{r pressure-stats}
pres_stats <- pres_long |>
  group_by(node) |>
  summarise(
    n          = n(),
    n_full_pay = sum(full_payload),
    mean_hPa   = mean(hPa, na.rm = TRUE),
    sd_hPa     = sd(hPa,   na.rm = TRUE),
    min_hPa    = min(hPa,  na.rm = TRUE),
    max_hPa    = max(hPa,  na.rm = TRUE),
    .groups    = "drop"
  )

pres_stats |>
  mutate(across(where(is.double), \(x) round(x, 2)),
         period = if_else(node <= 11, "full (Jan–Jun)", "old only (Jan–Mar)")) |>
  select(node, period, n, mean_hPa, sd_hPa, min_hPa, max_hPa) |>
  knitr::kable(
    col.names = c("Node", "Period", "N obs", "Mean (hPa)", "SD", "Min", "Max"),
    caption   = "Pressure statistics — all values within plausible range 600–800 hPa"
  )
```

```{r pressure-n1-variability}
# n1 vs n2-n11: is n1 genuinely more variable?
pres_n1_11 <- pres_long |> filter(node <= 11) |>
  mutate(group = if_else(node == 1, "n1 (surface)", "n2-n11 (subsurface)"))

ggplot(pres_n1_11, aes(time_utc, hPa, colour = group, group = node)) +
  geom_line(aes(alpha = group), linewidth = 0.7) +
  scale_colour_manual(values = c("n1 (surface)" = "firebrick",
                                  "n2-n11 (subsurface)" = "steelblue3"), name = NULL) +
  scale_alpha_manual(values = c("n1 (surface)" = 1, "n2-n11 (subsurface)" = 0.5),
                     guide = "none") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(title = "n1 vs n2–n11 pressure (full period)",
       subtitle = sprintf("n1: SD = %.1f hPa  |  n2-n11: SD = %.1f – %.1f hPa",
         pres_stats$sd_hPa[pres_stats$node == 1],
         min(pres_stats$sd_hPa[pres_stats$node %in% 2:11]),
         max(pres_stats$sd_hPa[pres_stats$node %in% 2:11])),
       x = NULL, y = "Pressure (hPa)") +
  theme_minimal(base_size = 13)
```

### Pressure variability vs node number

```{r pressure-variability-plot}
p_sd <- ggplot(pres_stats, aes(node, sd_hPa)) +
  geom_col(aes(fill = sd_hPa), show.legend = FALSE) +
  geom_text(aes(label = round(sd_hPa, 1)), vjust = -0.4, size = 2.8) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  labs(title = "Pressure variability (SD) per node",
       subtitle = "n1 clearly more variable — surface atmospheric influence",
       x = "Node", y = "SD (hPa)") +
  theme_minimal(base_size = 13)

p_mean <- ggplot(pres_stats, aes(node, mean_hPa)) +
  geom_point(aes(colour = sd_hPa), size = 3) +
  geom_line(colour = "grey60", linewidth = 0.5) +
  scale_colour_viridis_c(option = "plasma", direction = -1, name = "SD") +
  labs(title = "Mean pressure per node",
       subtitle = "Small offset between nodes — plausible depth/position differences",
       x = "Node", y = "Mean pressure (hPa)") +
  theme_minimal(base_size = 13)

p_sd | p_mean
```

### Seasonal pressure trend

```{r pressure-trend}
# Fit linear trend per node (n1-n11 full period)
pres_trends <- pres_long |>
  filter(node <= 11) |>
  mutate(t_days = as.numeric(difftime(time_utc, min(time_utc), units = "days"))) |>
  group_by(node) |>
  summarise(
    slope_hPa_month = coef(lm(hPa ~ t_days))[2] * 30,
    pval            = summary(lm(hPa ~ t_days))$coefficients[2, 4],
    .groups = "drop"
  ) |>
  mutate(sig = pval < 0.05)

ggplot(pres_trends, aes(node, slope_hPa_month, fill = sig)) +
  geom_col() +
  geom_hline(yintercept = 0, colour = "grey40") +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "steelblue3"),
                    labels = c("n.s.", "p < 0.05"), name = NULL) +
  labs(title = "Seasonal pressure trend per node (n1–n11, full period)",
       subtitle = "Positive = pressure increasing Jan → Jun (winter densification / seasonal atm. pressure)",
       x = "Node", y = "Trend (hPa / month)") +
  theme_minimal(base_size = 13)
```

---

## Battery voltage

Battery data is only available for the **old period** (untruncated payloads, Jan–Mar 2026).

```{r battery}
batt <- d |> filter(!is.na(battery_mv)) |>
  mutate(battery_V = battery_mv / 1000)

cat("battery_mv range:", range(d$battery_mv, na.rm=TRUE), "\n")

ggplot(batt, aes(time_utc, battery_V)) +
  geom_line(colour = "darkorange", linewidth = 0.9) +
  geom_point(colour = "darkorange", size = 1.5, alpha = 0.6) +
  scale_x_datetime(date_breaks = "1 week", date_labels = "%d %b") +
  labs(title = "Battery voltage (old period only — Jan 18 – Mar 5)",
       subtitle = "Values in mV from raw decode; divide by 1000 for V",
       x = NULL, y = "Voltage (V)") +
  theme_minimal(base_size = 13)
```

---

## Data completeness summary

```{r completeness}
# Show which nodes have data in which period
ntc_all <- grep("\\.ntc1_temp_C$", names(d), value=TRUE)
completeness <- tibble(
  node = as.integer(sub("n(\\d+)\\..*","\\1", ntc_all)),
  n_old = sapply(ntc_all, function(c)
    sum(!is.na(d[[c]]) & d$time_utc < CUTOFF_FULL)),
  n_new = sapply(ntc_all, function(c)
    sum(!is.na(d[[c]]) & d$time_utc >= CUTOFF_FULL))
) |>
  mutate(
    has_pressure = node %in% as.integer(sub("n(\\d+)\\..*","\\1", p_cols)),
    status = case_when(
      n_old > 0 & n_new > 0 ~ "full period",
      n_old > 0 & n_new == 0 ~ "old period only",
      TRUE ~ "no data"
    )
  )

completeness |>
  filter(n_old > 0 | n_new > 0) |>
  knitr::kable(
    col.names = c("Node", "N (Jan–5 Mar)", "N (6 Mar–now)", "Has pressure", "Coverage"),
    caption   = "Data availability per node. Both periods carry full payloads (all nodes) — old Rockblock export before 6 Mar, Cloudloop API after."
  )
```

---

## Notes

- **n1** = atmospheric temperature sensor (most variable; the red line in all plots)
- **All nodes n1–n35 valid across the whole record** — the data now come from the
  Cloudloop **API** (full 340 + 54 byte payload). The earlier n12+ artefacts were an
  artefact of the 128-B **web-export** truncation, which no longer applies.
- **battery_mv** available throughout; range `r paste(range(d$battery_mv, na.rm=TRUE), collapse="–")` mV
- Pressure increase Jan → Jun (~+15 hPa for n2–n11) is consistent with known seasonal
  atmospheric pressure cycle at Kohnen and possible firn compaction signal
