---
title: "head04 — Borehole Temperature Profile (Kohnen 2026)"
subtitle: "Modem 231710 | Sensors at 5–200 m depth | Decoded from Rockblock SBD"
author: "Thom Laepple"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 11
    fig-height: 6
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(patchwork)

d <- read_csv("./data/decoded_head04_231710_combined.csv", show_col_types = FALSE) |>
  mutate(
    time_utc = as_datetime(time_utc),
    t_days   = as.numeric(difftime(time_utc, min(time_utc), units = "days"))
  )
```

## Overview

Modem **231710** (head04) transmits once daily (~17:38 UTC). The chain carries
NTC thermistors at 5–200 m depth in the firn/ice at Kohnen station, Antarctica.
Due to a 128-byte payload truncation in the Rockblock CSV export, only nodes
**n1–n11** are reliably decoded from the current dataset; nodes n12–n25 require
the full payload via the Cloudloop API.

**Data range:** `r format(min(d$time_utc), "%d %b %Y")` – `r format(max(d$time_utc), "%d %b %Y")`
(`r nrow(d)` daily observations)

```{r node-metadata}
# Placeholder depths: evenly spaced 5–200 m across 25 total nodes
# Adjust node_depths once true installation depths are available
depth_step <- (200 - 5) / (25 - 1)
node_depths <- tibble(
  node  = 1:11,
  depth = round(5 + (0:10) * depth_step, 1)
)
knitr::kable(node_depths, col.names = c("Node", "Depth (m) [placeholder]"),
             caption = "Assumed sensor depths — replace with true values")
```

---

## Temperature time series

```{r timeseries}
ntc_long <- d |>
  select(time_utc, matches("^n[0-9]+\\.ntc1_temp_C$")) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  left_join(node_depths, by = "node")

ggplot(ntc_long, aes(time_utc, temp_C, colour = depth, group = node)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_c(option = "plasma", direction = -1,
                         name = "Tiefe (m)\n[Platzhalter]") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "NTC1 temperatures by depth — head04 (modem 231710)",
    subtitle = "Shallow nodes (n1, n2): strong seasonal signal  |  Deep nodes (n9–n11): quasi-constant",
    x = NULL, y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

The near-surface nodes (n1, n2) cool rapidly as the Antarctic winter sets in.
Below ~21 m (n3 onward) the amplitude decreases sharply and the signal becomes
quasi-stationary, reflecting the thermal diffusion length-scale in firn.

---

## Depth profiles at selected dates

```{r profiles}
profile_dates <- d |>
  pull(time_utc) |> as.Date() |> unique() |>
  (\(x) x[seq(1, length(x), length.out = 6)])() |>
  as.character()

profiles <- ntc_long |>
  mutate(date = as.character(as.Date(time_utc))) |>
  filter(date %in% profile_dates)

ggplot(profiles, aes(temp_C, depth, colour = date, group = date)) +
  geom_path(linewidth = 1, alpha = 0.9) +
  geom_point(size = 2.5) +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  scale_colour_viridis_d(option = "turbo", name = "Date") +
  labs(
    title    = "Temperature–depth profiles at selected dates",
    subtitle = "Cooling progresses from the surface downward; deep nodes remain stable",
    x = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

The profiles show the seasonal cooling wave propagating downward from March to
June. The profile shape is convex near the surface (strong gradient) and
asymptotically flat at depth, consistent with thermal diffusion in firn.

---

## Variability vs. depth

```{r variability}
node_stats <- ntc_long |>
  group_by(node, depth) |>
  summarise(
    mean_T  = mean(temp_C, na.rm = TRUE),
    sd_T    = sd(temp_C,   na.rm = TRUE),
    range_T = max(temp_C, na.rm = TRUE) - min(temp_C, na.rm = TRUE),
    .groups = "drop"
  )

p_sd <- ggplot(node_stats, aes(sd_T, depth)) +
  geom_path(colour = "grey60", linewidth = 0.6) +
  geom_point(aes(colour = sd_T), size = 4) +
  geom_text(aes(label = paste0("n", node)), hjust = -0.35, size = 3.2) +
  scale_colour_viridis_c(option = "plasma", direction = -1, guide = "none") +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  scale_x_log10() +
  labs(title = "Temperature variability vs. depth",
       subtitle = "Log scale: each additional ~8 m reduces SD by ~50%",
       x = "Standard deviation (°C, log scale)") +
  theme_minimal(base_size = 13)

p_mean <- ggplot(node_stats, aes(mean_T, depth)) +
  geom_path(colour = "grey60", linewidth = 0.6) +
  geom_point(aes(colour = depth), size = 4, show.legend = FALSE) +
  scale_colour_viridis_c(option = "plasma", direction = -1) +
  geom_text(aes(label = paste0("n", node)), hjust = -0.35, size = 3.2) +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  labs(title = "Mean temperature vs. depth",
       subtitle = sprintf("Deep mean annual temperature: ~%.2f°C (n11)",
                          node_stats$mean_T[node_stats$node == 11]),
       x = "Mean temperature (°C)") +
  theme_minimal(base_size = 13)

p_mean | p_sd
```

```{r variability-table}
node_stats |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(
    col.names = c("Node", "Depth (m)", "Mean T (°C)", "SD (°C)", "Range (°C)"),
    caption   = "Temperature statistics per node"
  )
```

---

## Linear trends and thermal diffusion

The 98-day record is long enough to reveal the phase structure of the seasonal
wave in the firn.

```{r trends}
ntc_cols <- grep("^n[0-9]+\\.ntc1_temp_C$", names(d), value = TRUE)

trends <- lapply(ntc_cols, function(col) {
  node <- as.integer(sub("n(\\d+)\\..*", "\\1", col))
  dat  <- data.frame(t = d$t_days, y = d[[col]]) |> na.omit()
  m    <- lm(y ~ t, data = dat)
  s    <- summary(m)
  tibble(node            = node,
         slope_mK_month  = coef(m)[2] * 30 * 1000,
         se_mK_month     = s$coefficients[2, 2] * 30 * 1000,
         pval            = s$coefficients[2, 4],
         mean_T          = mean(dat$y),
         sd_T            = sd(dat$y))
}) |>
  bind_rows() |>
  left_join(node_depths, by = "node") |>
  mutate(sig = pval < 0.05,
         sig_label = case_when(pval < 0.001 ~ "***",
                               pval < 0.01  ~ "**",
                               pval < 0.05  ~ "*",
                               TRUE         ~ "n.s."))

ggplot(trends, aes(slope_mK_month, depth)) +
  geom_vline(xintercept = 0, colour = "grey50", linetype = "dashed") +
  geom_errorbarh(aes(xmin = slope_mK_month - se_mK_month,
                     xmax = slope_mK_month + se_mK_month),
                 height = 2, colour = "grey60") +
  geom_point(aes(colour = sig, size = log10(abs(slope_mK_month) + 1))) +
  geom_text(aes(label = paste0("n", node, "  ", sig_label)),
            hjust = -0.15, size = 3.2) +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
                      labels = c("n.s. (p ≥ 0.05)", "p < 0.05"), name = NULL) +
  scale_size_continuous(range = c(2, 7), guide = "none") +
  labs(
    title    = "Linear trend per node (mK / month)",
    subtitle = "Negative = cooling (winter onset)  |  Positive = delayed summer diffusion",
    x        = "Trend (mK / month)"
  ) +
  theme_minimal(base_size = 13)
```

```{r trend-table}
trends |>
  mutate(across(c(slope_mK_month, se_mK_month, mean_T, sd_T), \(x) round(x, 2)),
         pval = formatC(pval, digits = 3, format = "g")) |>
  select(node, depth, mean_T, sd_T, slope_mK_month, se_mK_month, pval, sig_label) |>
  knitr::kable(
    col.names = c("Node", "Depth (m)", "Mean T", "SD", "Trend (mK/mon)", "SE", "p-value", ""),
    caption   = "Linear trend results. Positive trend in n4–n6 reflects phase-lagged downward diffusion of summer warmth."
  )
```

### Phase lag: mid-depth nodes warming while surface cools

```{r phase-lag}
mid_cols <- grep("^n[3456]\\.ntc1_temp_C$", names(d), value = TRUE)
d_mid <- d |>
  select(time_utc, all_of(mid_cols)) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  left_join(node_depths, by = "node")

ggplot(d_mid, aes(time_utc, temp_C, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.1, linewidth = 0.6) +
  scale_colour_brewer(palette = "Set1", name = "Node",
                      labels = paste0("n", 3:6)) +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "Nodes n3–n6: phase-lagged warming trend",
    subtitle = "Summer warmth still diffusing downward while the surface (n1, n2) cools toward winter",
    x = NULL, y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

### Deep nodes (n7–n11): quasi-constant

```{r deep-nodes}
deep_cols <- grep("^n([7-9]|1[01])\\.ntc1_temp_C$", names(d), value = TRUE)
d_deep <- d |>
  select(time_utc, all_of(deep_cols)) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  left_join(node_depths, by = "node")

ggplot(d_deep, aes(time_utc, temp_C, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed",
              linewidth = 0.4, alpha = 0.6) +
  scale_colour_viridis_d(option = "viridis", direction = -1,
                          name = "Node", labels = paste0("n", 7:11)) +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "Deep nodes n7–n11 (zoomed)",
    subtitle = sprintf("Total range < 0.1°C  |  n11 SD = %.4f°C  |  No significant trend detectable",
                       node_stats$sd_T[node_stats$node == 11]),
    x = NULL, y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

No statistically significant trend is detectable in nodes n7–n11 within the
98-day window. The estimated trends are < 5 mK/month with p > 0.05. Detecting
a geothermal baseline trend of ~0.1 mK/month would require at least ~1 year of
continuous data.

---

## testSB channel (calibration reference)

The `testSB` field is a firmware calibration reference — it should be constant
per node. The values (~6535–6542) show no temporal drift, confirming sensor
stability.

```{r testsb}
sb_long <- d |>
  select(time_utc, matches("testSB")) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "value") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor)))

ggplot(sb_long, aes(time_utc, value, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "testSB channel — calibration reference",
    subtitle = "Values constant over time per node → no sensor drift",
    x = NULL, y = "Value (counts / 1000)", colour = "Node"
  ) +
  theme_minimal(base_size = 13)
```

---

## Notes and caveats

- **Depth values are placeholders** — replace `node_depths` in the setup chunk
  with true installation depths once available.
- **Only n1–n11 decoded** — the Rockblock CSV export truncates payloads at
  128 bytes; full data (n1–n25 + battery_mv) requires the Cloudloop API.
- **Once-daily sampling** — one measurement per day (~17:38 UTC) gives no
  information on the diurnal cycle at depth, but at these depths the diurnal
  signal is negligible.
