---
title: "head04 — Borehole Temperature Profile (Kohnen 2026)"
subtitle: "Modem 231710 | Sensors at 5–200 m depth | Decoded from Rockblock SBD"
author: "Thom Laepple"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    fig-width: 11
    fig-height: 6
    theme: flatly
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(lubridate)
library(patchwork)

d <- read_csv("./data/decoded_head04_231710_combined.csv", show_col_types = FALSE) |>
  mutate(
    time_utc = as_datetime(time_utc),
    t_days   = as.numeric(difftime(time_utc, min(time_utc), units = "days"))
  )
```

## Overview

Modem **231710** (head04) transmits once daily (~17:38 UTC). The chain carries
NTC thermistors at 5–200 m depth in the firn/ice at Kohnen station, Antarctica.
All **n1–n25** are decoded: the data come from the **Cloudloop API**
(`fetch_antarctic_cloudloop.R`), which returns the full 284-byte payload. (The
earlier n1–n11 limit was an artefact of the 128-byte **web-export** truncation.)

**Data range:** `r format(min(d$time_utc), "%d %b %Y")` – `r format(max(d$time_utc), "%d %b %Y")`
(`r nrow(d)` daily observations)

```{r node-metadata}
# Placeholder depths: evenly spaced 5–200 m across all 25 nodes.
# (All 25 now decode via the Cloudloop API; under the old 128-B truncation only
# n1–n11 were available, so this table used to stop at n11.)
# Adjust node_depths once true installation depths are available.
depth_step <- (200 - 5) / (25 - 1)
node_depths <- tibble(
  node  = 1:25,
  depth = round(5 + (0:24) * depth_step, 1)
)
knitr::kable(node_depths, col.names = c("Node", "Depth (m) [placeholder]"),
             caption = "Assumed sensor depths — replace with true values")
```

---

## Temperature time series

```{r timeseries}
ntc_long <- d |>
  select(time_utc, matches("^n[0-9]+\\.ntc1_temp_C$")) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  left_join(node_depths, by = "node")

ggplot(ntc_long, aes(time_utc, temp_C, colour = depth, group = node)) +
  geom_line(linewidth = 0.8) +
  scale_colour_viridis_c(option = "plasma", direction = -1,
                         name = "Tiefe (m)\n[Platzhalter]") +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "NTC1 temperatures by depth — head04 (modem 231710)",
    subtitle = "Shallow nodes (n1, n2): strong seasonal signal  |  Deep nodes (n9–n11): quasi-constant",
    x = NULL, y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

The near-surface nodes (n1, n2) cool rapidly as the Antarctic winter sets in.
Below ~21 m (n3 onward) the amplitude decreases sharply and the signal becomes
quasi-stationary, reflecting the thermal diffusion length-scale in firn.

---

## Depth profiles at selected dates

```{r profiles}
profile_dates <- d |>
  pull(time_utc) |> as.Date() |> unique() |>
  (\(x) x[seq(1, length(x), length.out = 6)])() |>
  as.character()

profiles <- ntc_long |>
  mutate(date = as.character(as.Date(time_utc))) |>
  filter(date %in% profile_dates)

ggplot(profiles, aes(temp_C, depth, colour = date, group = date)) +
  geom_path(linewidth = 1, alpha = 0.9) +
  geom_point(size = 2.5) +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  scale_colour_viridis_d(option = "turbo", name = "Date") +
  labs(
    title    = "Temperature–depth profiles at selected dates",
    subtitle = "Cooling progresses from the surface downward; deep nodes remain stable",
    x = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

The profiles show the seasonal cooling wave propagating downward from March to
June. The profile shape is convex near the surface (strong gradient) and
asymptotically flat at depth, consistent with thermal diffusion in firn.

---

## Variability vs. depth

```{r variability}
node_stats <- ntc_long |>
  group_by(node, depth) |>
  summarise(
    mean_T  = mean(temp_C, na.rm = TRUE),
    sd_T    = sd(temp_C,   na.rm = TRUE),
    range_T = max(temp_C, na.rm = TRUE) - min(temp_C, na.rm = TRUE),
    .groups = "drop"
  )

p_sd <- ggplot(node_stats, aes(sd_T, depth)) +
  geom_path(colour = "grey60", linewidth = 0.6) +
  geom_point(aes(colour = sd_T), size = 4) +
  geom_text(aes(label = paste0("n", node)), hjust = -0.35, size = 3.2) +
  scale_colour_viridis_c(option = "plasma", direction = -1, guide = "none") +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  scale_x_log10() +
  labs(title = "Temperature variability vs. depth",
       subtitle = "Log scale: each additional ~8 m reduces SD by ~50%",
       x = "Standard deviation (°C, log scale)") +
  theme_minimal(base_size = 13)

p_mean <- ggplot(node_stats, aes(mean_T, depth)) +
  geom_path(colour = "grey60", linewidth = 0.6) +
  geom_point(aes(colour = depth), size = 4, show.legend = FALSE) +
  scale_colour_viridis_c(option = "plasma", direction = -1) +
  geom_text(aes(label = paste0("n", node)), hjust = -0.35, size = 3.2) +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  labs(title = "Mean temperature vs. depth",
       subtitle = sprintf("Deep mean annual temperature: ~%.2f°C (n11)",
                          node_stats$mean_T[node_stats$node == 11]),
       x = "Mean temperature (°C)") +
  theme_minimal(base_size = 13)

p_mean | p_sd
```

```{r variability-table}
node_stats |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(
    col.names = c("Node", "Depth (m)", "Mean T (°C)", "SD (°C)", "Range (°C)"),
    caption   = "Temperature statistics per node"
  )
```

---

## Linear trends and thermal diffusion

The `r round(max(d$t_days))`-day record (now full-length via the Cloudloop API)
is long enough to reveal the phase structure of the seasonal wave in the firn —
and, at the deep nodes, weak but statistically significant trends that the earlier
short record could not resolve.

```{r trends}
ntc_cols <- grep("^n[0-9]+\\.ntc1_temp_C$", names(d), value = TRUE)

trends <- lapply(ntc_cols, function(col) {
  node <- as.integer(sub("n(\\d+)\\..*", "\\1", col))
  dat  <- data.frame(t = d$t_days, y = d[[col]]) |> na.omit()
  m    <- lm(y ~ t, data = dat)
  s    <- summary(m)
  tibble(node            = node,
         slope_mK_month  = coef(m)[2] * 30 * 1000,
         se_mK_month     = s$coefficients[2, 2] * 30 * 1000,
         pval            = s$coefficients[2, 4],
         mean_T          = mean(dat$y),
         sd_T            = sd(dat$y))
}) |>
  bind_rows() |>
  left_join(node_depths, by = "node") |>
  mutate(sig = pval < 0.05,
         sig_label = case_when(pval < 0.001 ~ "***",
                               pval < 0.01  ~ "**",
                               pval < 0.05  ~ "*",
                               TRUE         ~ "n.s."))

ggplot(trends, aes(slope_mK_month, depth)) +
  geom_vline(xintercept = 0, colour = "grey50", linetype = "dashed") +
  geom_errorbarh(aes(xmin = slope_mK_month - se_mK_month,
                     xmax = slope_mK_month + se_mK_month),
                 height = 2, colour = "grey60") +
  geom_point(aes(colour = sig, size = log10(abs(slope_mK_month) + 1))) +
  geom_text(aes(label = paste0("n", node, "  ", sig_label)),
            hjust = -0.15, size = 3.2) +
  scale_y_reverse(name = "Depth (m) [placeholder]") +
  scale_colour_manual(values = c("FALSE" = "grey60", "TRUE" = "firebrick"),
                      labels = c("n.s. (p ≥ 0.05)", "p < 0.05"), name = NULL) +
  scale_size_continuous(range = c(2, 7), guide = "none") +
  labs(
    title    = "Linear trend per node (mK / month)",
    subtitle = "Negative = cooling (winter onset)  |  Positive = delayed summer diffusion",
    x        = "Trend (mK / month)"
  ) +
  theme_minimal(base_size = 13)
```

```{r trend-table}
trends |>
  mutate(across(c(slope_mK_month, se_mK_month, mean_T, sd_T), \(x) round(x, 2)),
         pval = formatC(pval, digits = 3, format = "g")) |>
  select(node, depth, mean_T, sd_T, slope_mK_month, se_mK_month, pval, sig_label) |>
  knitr::kable(
    col.names = c("Node", "Depth (m)", "Mean T", "SD", "Trend (mK/mon)", "SE", "p-value", ""),
    caption   = "Linear trend results. Positive trend in n4–n6 reflects phase-lagged downward diffusion of summer warmth."
  )
```

### Phase lag: mid-depth nodes warming while surface cools

```{r phase-lag}
mid_cols <- grep("^n[3456]\\.ntc1_temp_C$", names(d), value = TRUE)
d_mid <- d |>
  select(time_utc, all_of(mid_cols)) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  left_join(node_depths, by = "node")

ggplot(d_mid, aes(time_utc, temp_C, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.1, linewidth = 0.6) +
  scale_colour_brewer(palette = "Set1", name = "Node",
                      labels = paste0("n", 3:6)) +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "Nodes n3–n6: phase-lagged warming trend",
    subtitle = "Summer warmth still diffusing downward while the surface (n1, n2) cools toward winter",
    x = NULL, y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

### Deep nodes (n7–n11): trend over the full record

Zoomed to the deep, near-constant nodes. Over the **full `r round(max(d$t_days))`-day
record** the deep-node trend (dashed = linear fit) is re-evaluated per node — the
earlier 98-day window was too short to separate any trend from noise.

```{r deep-nodes}
deep_nodes <- 7:11
deep_cols  <- sprintf("n%d.ntc1_temp_C", deep_nodes)

# Per-node linear fit over the WHOLE record (mK/month + significance).
deep_trends <- lapply(deep_nodes, function(n) {
  dat <- na.omit(data.frame(t = d$t_days, y = d[[sprintf("n%d.ntc1_temp_C", n)]]))
  s   <- summary(lm(y ~ t, dat))
  tibble(node = n, slope_mK_month = s$coefficients[2, 1] * 30 * 1000,
         se = s$coefficients[2, 2] * 30 * 1000, pval = s$coefficients[2, 4],
         sd_T = sd(dat$y))
}) |> bind_rows()
n_sig <- sum(deep_trends$pval < 0.05)

d_deep <- d |>
  select(time_utc, all_of(deep_cols)) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "temp_C") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor))) |>
  left_join(node_depths, by = "node")

ggplot(d_deep, aes(time_utc, temp_C, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.8) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed",
              linewidth = 0.4, alpha = 0.6) +
  scale_colour_viridis_d(option = "viridis", direction = -1,
                          name = "Node", labels = paste0("n", deep_nodes)) +
  scale_x_datetime(date_breaks = "1 month", date_labels = "%d %b") +
  labs(
    title    = "Deep nodes n7–n11 — full record",
    subtitle = sprintf("n11 SD = %.4f°C  |  %d of 5 deep nodes show a significant trend (p<0.05) over %d days",
                       deep_trends$sd_T[deep_trends$node == 11], n_sig, round(max(d$t_days))),
    x = NULL, y = "Temperature (°C)"
  ) +
  theme_minimal(base_size = 13)
```

```{r deep-trend-table}
deep_trends |>
  mutate(across(c(slope_mK_month, se, sd_T), \(x) round(x, 3)),
         pval = formatC(pval, digits = 2, format = "g"),
         sig  = ifelse(pval < 0.05, "yes", "—")) |>
  knitr::kable(col.names = c("Node", "Trend (mK/mon)", "SE", "p-value", "SD (°C)", "p<0.05"),
    caption = "Deep-node linear trend over the full record.")
```

Unlike the earlier 98-day window (where no deep-node trend was resolvable), the
full `r round(max(d$t_days))`-day record reveals **weak but significant warming**
at the deepest nodes — **n9 ≈ `r sprintf("%+.1f", deep_trends$slope_mK_month[deep_trends$node==9])`**,
**n10 ≈ `r sprintf("%+.1f", deep_trends$slope_mK_month[deep_trends$node==10])`** and
**n11 ≈ `r sprintf("%+.1f", deep_trends$slope_mK_month[deep_trends$node==11])` mK/month**
(all p < 0.05). The sign and downward-decreasing magnitude are consistent with the
**phase-lagged summer wave still diffusing to depth**, not a geothermal baseline
(which would be ~0.1 mK/month and needs multi-year data). n11 remains the most
stable sensor (SD `r sprintf("%.4f", deep_trends$sd_T[deep_trends$node==11])` °C).

---

## testSB channel (calibration reference)

The `testSB` field is a firmware calibration reference — it should be constant
per node. The values (~6535–6542) show no temporal drift, confirming sensor
stability.

```{r testsb}
sb_long <- d |>
  select(time_utc, matches("testSB")) |>
  pivot_longer(-time_utc, names_to = "sensor", values_to = "value") |>
  mutate(node = as.integer(sub("n(\\d+)\\..*", "\\1", sensor)))

ggplot(sb_long, aes(time_utc, value, colour = factor(node), group = node)) +
  geom_line(linewidth = 0.5, alpha = 0.8) +
  scale_x_datetime(date_breaks = "2 weeks", date_labels = "%d %b") +
  labs(
    title    = "testSB channel — calibration reference",
    subtitle = "Values constant over time per node → no sensor drift",
    x = NULL, y = "Value (counts / 1000)", colour = "Node"
  ) +
  theme_minimal(base_size = 13)
```

---

## Notes and caveats

- **Depth values are placeholders** — replace `node_depths` in the setup chunk
  with true installation depths once available.
- **All n1–n25 + battery decoded** — via the Cloudloop API (full 284-byte payload);
  the former n1–n11 limit was the 128-byte web-export truncation, now resolved.
- **Once-daily sampling** — one measurement per day (~17:38 UTC) gives no
  information on the diurnal cycle at depth, but at these depths the diurnal
  signal is negligible.
